% ==============================================================================
% Reframed Paper -- Methods-First Version
% Target: Journal of Archaeological Science
%
% COMPILE INSTRUCTIONS:
%   Place this file in the same directory as generated_macros.tex, then run:
%     pdflatex reframed_paper.tex
%   Or adjust the \input path below to match your project layout.
% ==============================================================================

\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{array}
\usepackage{natbib}
\usepackage{url}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{hyperref}
\setlength{\emergencystretch}{3em}
\newcommand{\degr}{\textdegree{}}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!70!black,
  citecolor=blue!70!black,
  urlcolor=blue!60!black
}

\input{generated_macros}

\begin{document}

\title{The Wrong Test for the Right Question:\\
Geometric-Null Binomial Proximity Testing\\
for Longitude Quantization in Monument Distributions}

\author{Seth Tenenbaum\thanks{ORCID:
\href{https://orcid.org/0009-0008-5797-2498}{0009-0008-5797-2498}.
Correspondence: seth@fourthtemple.com}\\
Independent Scholar, USA}

\date{May 2026}

\maketitle

\begin{abstract}
Claims that ancient or sacred monuments align with longitude grids
have appeared repeatedly in the archaeoastronomical literature and been
repeatedly criticised --- but almost always on grounds of
investigator-assembled site lists rather than statistical framework.
This paper identifies a more fundamental problem: the Rayleigh
periodogram, the field's standard circular-statistical tool, is
structurally incapable of testing the hypothesis it is applied to.
The hypothesis ``sites cluster near specific harmonic longitude lines''
is a proximity question.
The Rayleigh statistic answers a different question: whether all site
phases share a single preferred direction.
A true grid-placement signal distributes sites near \emph{multiple}
harmonic lines simultaneously, producing \emph{low} Rayleigh $R$,
not high.
We demonstrate this empirically: a period sweep of the UNESCO
context-validated dome/stupa corpus ($N = \circTwoSampNDome{}$) shows
Rayleigh significance at arbitrary periods ($2.4\degr$, $3.5\degr$,
$3.6\degr$) with no corresponding harmonic proximity enrichment, while
the only period with genuine proximity enrichment ($3.0\degr$,
$p_\mathrm{binom} = \sweepBinomPThreeZero{}$) is Rayleigh-non-significant.
The correct test is a one-sided binomial on site counts within a
threshold distance of any harmonic, against the geometric null
$p_0 = 2d/T$.
We apply this framework to three independently assembled hemispherical
monument corpora --- UNESCO World Heritage dome/stupa sites ($N =
\circTwoSampNDome{}$ context-validated), Wikidata Q180987 stupas
($N = \npcNStupa{}$), and a crowd-sourced OpenStreetMap stupa layer
($N = \osmStupaTotal{}$) --- as a worked demonstration.
All three show directional excess over the geometric null at the
$3\degr$ / ${\sim}35\degr$E candidate frame; their independent
binomial p-values combine via Fisher's method to $p = \geoFisherAPDisp{}$ at the
broadest tier.
This result is offered as illustration, not proof of a historical grid:
corpus-frame conditioning, geographic dependence, and the mechanism gap
are fully acknowledged.
The methodological contribution --- the dissociation between Rayleigh
periodicity and harmonic proximity, and the geometric-null binomial
framework that correctly tests the latter --- applies to any claim of
this type regardless of context.
\end{abstract}

\bigskip
\noindent\textbf{Keywords:}
spatial statistics, circular statistics, Rayleigh test, harmonic
proximity, binomial geometric null, monument distribution, longitude
quantization, stupa, UNESCO World Heritage, archaeoastronomy

\bigskip

\section{Introduction}
\label{sec:introduction}

Do monuments cluster near specific longitude intervals? The question
recurs in the archaeoastronomical and archaeogeographic literature,
usually motivated by observational coincidences involving sacred sites,
trade corridors, or cosmographic traditions. It has also been repeatedly
attacked --- but almost always on grounds of investigator-assembled site
lists, insufficient correction for multiple comparisons, or lack of
independent replication
\citep{williamson1983leylines,freeman1976,kendall1974}.
These are legitimate criticisms, but they leave a more fundamental
problem unaddressed: the standard statistical tool applied to this class
of question is structurally incapable of testing it.

The tool in question is the Rayleigh statistic $R$. $R$ measures the
degree to which a set of phases concentrates toward a single preferred
direction on the unit circle --- it detects unimodal directional
concentration. Prior studies have applied Rayleigh tests to longitude
distributions and interpreted a significant $R$ at some period $T$ as
evidence that sites are placed near the harmonics of $T$
\citep{thom1967,magli2013,magli2016}.

This interpretation is wrong. A grid-placement hypothesis predicts that
sites fall near \emph{multiple} harmonic lines at spacing $T$ --- near
longitudes $\lambda_0, \lambda_0 + T, \lambda_0 + 2T$, and so on.
That distributes phases near $0, 2\pi/K, 4\pi/K, \ldots$ on the unit
circle, which is precisely the pattern that produces a \emph{low} $R$,
not a high one. Rayleigh significance at period $T$ is evidence that
all sites share one phase --- a single preferred longitude modulo $T$
--- which is neither what grid-placement predicts nor what the
hypothesis requires. The two statistics address orthogonal questions.

The correct test for ``sites cluster near harmonics'' is
straightforward. Under a uniform longitude distribution, the fraction
of sites falling within $\pm d\degr$ of any harmonic of period $T$
is exactly $p_0 = 2d/T$, regardless of geographic clustering. This
is the \emph{geometric null}. The appropriate statistic is a one-sided
binomial test on the count of sites within threshold $d$ of any
harmonic, against expected $N \times p_0$. When multiple independently
assembled corpora are available, their binomial p-values combine via
Fisher's method to give a joint estimate not inflated by within-corpus
geographic structure, because the corpora's geographic clusters are
non-overlapping.

This paper makes the dissociation explicit, demonstrates it
empirically, and proposes the geometric-null binomial framework as the
correct inferential tool for this class of hypothesis. We illustrate
the framework using three independently assembled corpora of
hemispherical monuments as a worked example at a candidate $3\degr$ /
${\sim}35\degr$E frame. The empirical result is offered as illustration
rather than proof: we make no claim for a historical grid, intentional
placement, or unique origin.

\section{Why the Rayleigh Test Fails for Harmonic Proximity}
\label{sec:rayleigh_failure}

\subsection{The Structural Mismatch}

Let $\lambda_1, \ldots, \lambda_N$ be site longitudes and $T$ a
candidate grid period. The Rayleigh statistic maps each longitude to a
phase $\theta_i = 2\pi(\lambda_i \bmod T)/T$ and computes

\[
R = \left| \frac{1}{N} \sum_{i=1}^{N} e^{i\theta_i} \right|.
\]

$R$ is large when all $\theta_i$ concentrate near the same value:
when sites cluster near a \emph{single} longitude modulo $T$.
$R$ is small when $\theta_i$ are dispersed, including the case where
they concentrate at \emph{multiple} equally spaced values.

A grid-placement hypothesis predicts the latter. If a set of harmonic
longitudes $\{\lambda_0 + kT : k = 0,1,2,\ldots\}$ attracts monument
placement, sites will cluster near each harmonic separately. For any
geographically spread corpus with sites near several different
harmonics, $R$ will be small. High Rayleigh $R$ at period $T$ would
instead indicate that all sites fall near the \emph{same} harmonic ---
a much stronger and less plausible claim.

A significant Rayleigh $R$ at an arbitrary period $T$ in a
geographically clustered corpus can arise whenever one geographic
cluster dominates the phase distribution --- not because sites are near
harmonics, but because they are near one longitude and that longitude
produces a peaked distribution at the tested period. This is a generic
property of clustered data on the unit circle, and it fires at many
arbitrary periods that have no proximity interpretation.

\subsection{Empirical Demonstration: Period Sweep Dissociation}

Table~\ref{tab:rayleigh_vs_binom} demonstrates this concretely using
the UNESCO context-validated dome/stupa corpus ($N = \circTwoSampNDome{}$).
A period sweep tests both Rayleigh concentration and binomial proximity
enrichment at the A$++$ precision threshold ($\pm 0.10 \times T$,
geometric null $p_0 = 0.20$) at four candidate periods.

\begin{table}[ht]
\centering
\caption{Rayleigh concentration vs.\ A$++$ binomial proximity
enrichment at four candidate periods, UNESCO dome corpus
($N = \circTwoSampNDome{}$, anchor \GerizimLon\degr{}E).
A$++$ threshold $= 0.10 \times T$; geometric null $p_0 = 0.20$.
Periods with Rayleigh significance show no proximity enrichment; the
only period with genuine proximity enrichment ($3.0\degr$) is
Rayleigh-non-significant. The two statistics are anti-correlated.}
\label{tab:rayleigh_vs_binom}
\small
\begin{tabular}{rlcrrrl}
\toprule
Period & Rayleigh $p$ & Sig & A$++$ hits & Expected
       & Binom.\ $p$ & Sig \\
\midrule
$2.4\degr$ & \sweepRayleighPTwoFour{}  & \sweepRayleighSigTwoFour{}
           & \sweepHitsTwoFour{}  & \sweepExpectedTwoFour{}
           & \sweepBinomPTwoFour{}  & \sweepBinomSigTwoFour{} \\
$3.0\degr$ & \sweepRayleighPThreeZero{} & \sweepRayleighSigThreeZero{}
           & \sweepHitsThreeZero{} & \sweepExpectedThreeZero{}
           & \sweepBinomPThreeZero{} & \textbf{\sweepBinomSigThreeZero{}} \\
$3.5\degr$ & \sweepRayleighPThreeFive{} & \sweepRayleighSigThreeFive{}
           & \sweepHitsThreeFive{} & \sweepExpectedThreeFive{}
           & \sweepBinomPThreeFive{} & \sweepBinomSigThreeFive{} \\
$3.6\degr$ & \sweepRayleighPThreeSix{}  & \sweepRayleighSigThreeSix{}
           & \sweepHitsThreeSix{}  & \sweepExpectedThreeSix{}
           & \sweepBinomPThreeSix{}  & \sweepBinomSigThreeSix{} \\
\bottomrule
\end{tabular}
\end{table}

The anti-correlation is unambiguous. Rayleigh is significant at
$2.4\degr$, $3.5\degr$, and $3.6\degr$, and non-significant at
$3.0\degr$. The binomial proximity test is significant only at
$3.0\degr$. A researcher applying the Rayleigh test would identify
the wrong periods and miss the one with actual proximity structure.

The geometric explanation is direct. At $2.4\degr$, $3.5\degr$, and
$3.6\degr$, one geographic cluster dominates the phase distribution,
producing spuriously peaked $R$. At $3.0\degr$, the cluster structure
spreads across multiple harmonics, lowering $R$ while the binomial
count of sites near any harmonic is elevated. The Rayleigh test
detects geographic clustering unrelated to harmonic alignment; the
binomial test detects the alignment directly.

\section{The Geometric-Null Binomial Framework}
\label{sec:framework}

\subsection{The Geometric Null}

For a grid of period $T$ anchored at $\lambda_0$, the deviation of
site $i$ from its nearest harmonic is

\[
\delta_i = \min_{k \in \mathbb{Z}} |\lambda_i - (\lambda_0 + kT)|,
\quad \text{folded to } [0, T/2].
\]

A site falls within tier threshold $d$ if $\delta_i \leq d$. Under a
uniform longitude distribution, $\Pr(\delta_i \leq d) = 2d/T$
regardless of geographic clustering. Clustering is \emph{neutral}:
a dense cluster anywhere inflates both observed and expected hit counts
proportionally, contributing to the binomial test only if the cluster
falls asymmetrically near harmonic lines. Geographic concentration is
therefore not a confound to be removed, but a source of statistical
power that the test uses correctly.

\subsection{Tier Classification}

We define three proximity thresholds as round fractions of the $T/2$
half-period ($1.5\degr$), calibrated to bracket the empirical natural
proximity scales of the corpora under study: the OWTRAD Silk Road
network (2.6\%), the UNESCO dome subset (3.6\%), the full UNESCO
Cultural/Mixed corpus (10.2\%), and the Wikidata stupa corpus (20.2\%).
Each threshold corresponds to a geometric null rate $p_0 = 2d/T$:

\begin{center}
\small
\begin{tabular}{lll}
\toprule
Tier & Threshold & Geometric null $p_0$ \\
\midrule
A$++$ & $\delta \leq \AppThreshDeg\degr$ ($\lesssim \AppThreshKm$~km equatorial) & \NullRateApp \\
A$+$  & $\delta \leq \ApThreshDeg\degr$  ($\lesssim \ApThreshKm$~km)             & \NullRateAp  \\
A     & $\delta \leq \ATierThreshDeg\degr$ ($\lesssim \ATierThreshKm$~km)        & \NullRateA   \\
\bottomrule
\end{tabular}
\end{center}

Tier labels are descriptive shorthand; the primary test is always the
binomial.

\subsection{Cross-Corpus Fisher Combination}

For each corpus $c$ at tier $d$, a one-sided binomial test yields
$p_c$. Fisher's combination statistic $\chi^2 = -2\sum_c \ln p_c$,
referred to $\chi^2(2C)$ for $C$ corpora, tests whether the joint
directional excess is consistent with chance across all corpora
simultaneously. The geographic clusters of the three corpora are
substantially non-overlapping, so the joint result cannot be
attributed to a single shared geographic cluster. This three-corpus
Fisher combination is the primary statistical result of the paper.
Rayleigh statistics are retained for comparability with prior
literature only.

\section{Data}
\label{sec:data}

\textbf{UNESCO dome/stupa subset ($N = \circTwoSampNDome{}$).}
The UNESCO World Heritage XML dataset \citep{unescowhc2024xml} contains
\NclusterTotal{} Cultural and Mixed inscribed sites. Seven morphological
keywords (stupa/stupas, tholos, dome/domed/domes, spherical) were
matched against site names, XML descriptions, and OUV text. Context
validation reduces the raw match of $N = \NcircTotal{}$ to
$N = \NcircValidated{}$, applying sentence-level gating to ambiguous
terms while accepting unambiguous terms (stupa, tholos) without gating.
These results are conditional: the $3\degr$ / ${\sim}35\degr$E frame
was identified in this corpus first, so UNESCO results cannot
independently confirm the frame.

\textbf{Wikidata Q180987 stupa corpus ($N = \npcNStupa{}$).}
All \textit{instance-of: stupa} (Q180987) items in the Wikidata
knowledge base, extracted without researcher involvement. Membership
was fixed before any statistics were computed. The corpus is
geographically dominated by the Java/Sumatra node
(${\sim}105$--$115\degr$E), reflecting the documented historical density
of stupa construction around Borobudur \citep{miksic1990borobudur}.
This corpus carries no look-elsewhere conditioning.

\textbf{OSM stupa layer ($N = \osmStupaTotal{}$).}
OpenStreetMap features with tags \texttt{building=stupa},
\texttt{historic=stupa}, or \texttt{ruins=stupa}, assembled by
crowd-sourced tagging independent of both UNESCO and Wikidata. Coverage
correlates with mapping effort; the corpus is treated as corroborating
evidence. Uneven contributor density is a coverage limitation, not a
directional bias: it would produce an arbitrary anchor, not the same
${\sim}35\degr$E line recovered by all three corpora.

\textbf{Candidate frame.}
The reference anchor is Mount Gerizim (\GerizimLon\degr{}E), a
prominent peak in the central Levantine highlands at the geographic
junction between the Mediterranean coast and the interior corridor
running toward Mesopotamia and beyond. It is not inscribed on the
UNESCO World Heritage List. Its longitude falls within the candidate
band identified in the data, and it serves here as a named label for
that corridor-level phase rather than as a unique or privileged point.
The anchor is band-invariant: every longitude in the \xbandLabel\degr{}E
band produces the same proximity enrichment, and both Gerizim and
Jerusalem (\JerusalemLon\degr{}E, $\approx \GerizimJerusalemKm$~km
south) place the corridor in the global top \anchorSweepTopPct\% of
\anchorSweepNanchors{} trial anchors. The data cannot distinguish
Gerizim from any co-grid longitude within the same A$+$ tier.

\section{Results}
\label{sec:results}

\subsection{Primary Result: Cross-Corpus Geometric-Null Test}

Table~\ref{tab:cross_corpus_geo} reports the primary statistical
results: one-sided binomial tests and Fisher combinations across all
three corpora.

\begin{table}[ht]
\centering
\caption{Cross-corpus geometric-null proximity test. One-sided
binomial: observed harmonic-proximity hits vs.\ expected under
$p_0 = 2d/T$. Fisher combination: $\chi^2(-2\sum\ln p_i)$,
$\chi^2(6)$, three corpora. UNESCO results are conditional; Wikidata
and OSM results carry no such conditioning.}
\label{tab:cross_corpus_geo}
\small
\begin{tabular}{llrrrrrl}
\toprule
Tier & Corpus & $N$ & Hits & Exp & Fold & $p_\mathrm{binom}$ & Sig \\
\midrule
\multirow{4}{*}{A$++$ ($p_0{=}\NullRateApp{}$)}
  & UNESCO domes    & \geoDomeAppN{} & \geoDomeAppHits{} & \geoDomeAppExp{} & \geoDomeAppEnr{}$\times$ & \geoDomeAppP{} & \geoDomeAppSig{} \\
  & Wikidata stupas & \geoWikiAppN{} & \geoWikiAppHits{} & \geoWikiAppExp{} & \geoWikiAppEnr{}$\times$ & \geoWikiAppP{} & \geoWikiAppSig{} \\
  & OSM stupas      & \geoOsmAppN{}  & \geoOsmAppHits{}  & \geoOsmAppExp{}  & \geoOsmAppEnr{}$\times$  & \geoOsmAppP{}  & \geoOsmAppSig{} \\
  & \multicolumn{6}{l}{\textit{Fisher $\chi^2(6) = \geoFisherAppChi{}$, combined $p = \geoFisherAppP{}$}} & \geoFisherAppSig{} \\
\midrule
\multirow{4}{*}{A$+$ ($p_0{=}\NullRateAp{}$)}
  & UNESCO domes    & \geoDomeApN{}  & \geoDomeApHits{}  & \geoDomeApExp{}  & \geoDomeApEnr{}$\times$  & \geoDomeApP{}  & \geoDomeApSig{} \\
  & Wikidata stupas & \geoWikiApN{}  & \geoWikiApHits{}  & \geoWikiApExp{}  & \geoWikiApEnr{}$\times$  & \geoWikiApP{}  & \geoWikiApSig{} \\
  & OSM stupas      & \geoOsmApN{}   & \geoOsmApHits{}   & \geoOsmApExp{}   & \geoOsmApEnr{}$\times$   & \geoOsmApP{}   & \geoOsmApSig{} \\
  & \multicolumn{6}{l}{\textit{Fisher $\chi^2(6) = \geoFisherApChi{}$, combined $p = \geoFisherApP{}$}} & \geoFisherApSig{} \\
\midrule
\multirow{4}{*}{A ($p_0{=}\NullRateA{}$)}
  & UNESCO domes    & \geoDomeAN{}   & \geoDomeAHits{}   & \geoDomeAExp{}   & \geoDomeAEnr{}$\times$   & \geoDomeAP{}   & \geoDomeASig{} \\
  & Wikidata stupas & \geoWikiAN{}   & \geoWikiAHits{}   & \geoWikiAExp{}   & \geoWikiAEnr{}$\times$   & \geoWikiAP{}   & \geoWikiASig{} \\
  & OSM stupas      & \geoOsmAN{}    & \geoOsmAHits{}    & \geoOsmAExp{}    & \geoOsmAEnr{}$\times$    & \geoOsmAP{}    & \geoOsmASig{} \\
  & \multicolumn{6}{l}{\textit{Fisher $\chi^2(6) = \geoFisherAChi{}$, combined $p = \geoFisherAP{}$}} & \geoFisherASig{} \\
\bottomrule
\end{tabular}
\end{table}

All three corpora show consistent directional excess at every tier
despite substantially different geographic footprints and assembly
processes. The Fisher combination reaches conventional significance at
all three tiers. The UNESCO A$++$ result is the strongest individual
signal ($\geoDomeAppEnr{}\times$), reflecting the distributed character
of that corpus; Wikidata and OSM A$++$ results are individually
non-significant, reflecting cluster-level rather than individual-site
precision. The Tier~A Fisher combination ($p = \geoFisherAP{}$) is the
most robust result.

\subsection{Supplementary Rayleigh Phase Concentration}

Table~\ref{tab:rayleigh_summary} reports Rayleigh statistics at
$T = 3\degr$ for comparability with prior literature. The full UNESCO
corpus shows no Rayleigh periodicity ($R = \circConcUnescoR{}$,
perm-$p = \circConcUnescoP{}$, ns), confirming the signal is not a
generic heritage-list property. Watson's $U^2$ and Kuiper tests confirm
the dome/stupa phase distribution differs detectably from the full
UNESCO corpus ($p = \circWatsonPUnescoStupa{}$,
$p = \circKuiperPUnescoStupa{}$), establishing monument-type
specificity consistent with the geometric-null results.

\begin{table}[ht]
\centering
\caption{Supplementary Rayleigh phase concentration at $T = 3\degr$,
retained for comparability with prior literature. The primary evidence
is Table~\ref{tab:cross_corpus_geo}.}
\label{tab:rayleigh_summary}
\small
\begin{tabular}{llrrrl}
\toprule
Corpus & $N$ & $R$ & perm-$p$ & Sig & Note \\
\midrule
UNESCO full corpus   & \NclusterTotal{} & \circConcUnescoR{} & \circConcUnescoP{} & ns & Boundary condition \\
UNESCO dome/stupa    & \circTwoSampNDome{} & \circConcDomeR{} & \circConcDomeP{} & \circConcDomePSig{} & Conditional \\
Wikidata Q180987     & \npcNStupa{} & \circConcStupaR{} & \circConcStupaP{} & \circConcStupaPSig{} & Independent \\
OSM stupa layer      & \osmStupaTotal{} & \osmStupaRayleighR{} & \osmStupaRayleighP{} & * & Crowd-sourced \\
\midrule
\multicolumn{6}{l}{\textit{Fisher (Wikidata + OSM): $\chi^2(4) = \osmStupaFisherTwoCorpusChi{}$, $p = \osmStupaFisherTwoCorpusP{}$}} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Negative Controls and Scope}

The OWTRAD Silk Road route network ($N = \npcNOwtrad{}$ nodes) does
not independently recover the monument anchor in the Normalized Phasor
Consensus test ($K = \npcNegUnOwK{}$, perm-$p = \npcNegUnOwPermP{}$,
ns). Using the \NOwtradEdges{} edge midpoints as the proximity sample
(one value per route segment, non-deduplicated), the binomial test is
also non-significant at both tiers (A$++$: $p = \pOwtradMidApp{}$,
ns; A$+$: $p = \pOwtradMidAp{}$, ns). The route network therefore
fails both the directional and the per-segment proximity tests.
By contrast, when route endpoints are pooled with multiplicity
($N = \NOwtradDegW{}$ node-in-edge occurrences, counting each
city once per edge it connects), the binomial enrichment is
significant (A$++$: $p = \pOwtradDegWApp{}$, $z = \zOwtradDegWApp{}$;
A$+$: $p = \pOwtradDegWAp{}$). The UNESCO Americas sub-corpus ($N = \npcNAmericas{}$) shows A$+$
rate of 9.7\%, below the \NullRateAp{} geometric null. A Monte Carlo
period sweep finds the dominant peak at $3\degr$ ($p < 0.001$); adjacent
non-harmonic spacings ($1.8\degr$, $2.1\degr$, $2.7\degr$, $3.3\degr$)
are non-significant in both populations.

Null~A (within-dome bootstrap, \simNperms{} trials) shows the UNESCO
A$++$ count is fully reproducible from dome/stupa sites' own longitude
distribution ($p = \geoNullDomeBootP{}$, ns): the enrichment relative
to the full corpus is carried by where hemispherical monuments are
located. Null~C finds the broader same-footprint UNESCO background
expects only ${\sim}\geoNullDomeRestrictedMeanfive{}$ A$++$ hits versus the
observed \NcircTierApp{} ($p = \geoNullDomeRestrictedPfive{}$ at the
${\pm}5\degr$ window, surviving Bonferroni correction for $k = 3$
windows, $\alpha_\mathrm{adj} = 0.0167$), establishing
type-specificity within that geography.

\subsection{Anchor Concordance}

All three corpora recover the same ${\sim}35\degr$E anchor. The
Normalized Phasor Consensus statistic $K$ --- which compares normalized
mean phasors at the candidate period, independent of tier counts ---
yields $K \geq \npcOsmStupaK{}$ for all pairings of the three monument
corpora (Table~\ref{tab:npc_key}). The three-corpus combination reaches
perm-$p = \npcOsmThreePermP{}$ with consensus anchor at
$\npcOsmThreeGridDeltaE\degr$ from $35\degr$E. Negative controls show
$K$ drops sharply when the monument corpora are replaced with
geographic or route-network controls.

\begin{table}[ht]
\centering
\caption{Selected NPC anchor-concordance configurations at
$T = 3\degr$ (\simNperms{} permutation draws). High $K$ with
non-significant two-corpus perm-$p$ reflects shared Eurasian geography.
Negative controls (Rows~4--5) confirm $K$ is not a generic Eurasian
property.}
\label{tab:npc_key}
\small
\begin{tabular}{llrrl}
\toprule
Role & Corpora & $K$ & perm-$p$ & Note \\
\midrule
Non-nested pair  & Dome + Wikidata Stupa      & \npcDomeStupaK{} & \npcDomeStupaPermP{} & Primary non-nested \\
Conservative     & UNESCO full + Stupa         & \npcTwoK{}        & \npcTwoPermP{}        & Non-nested \\
Three-corpus     & Dome + Stupa + OSM          & \npcOsmThreeK{}   & \npcOsmThreePermP{}   & All 3 monument corpora \\
Neg.\ control    & UNESCO + Americas + Stupa   & \npcNegThreeK{}   & \npcNegThreePermP{}   & Geographic control \\
Contrast corpus  & UNESCO + OWTRAD             & \npcNegUnOwK{}    & \npcNegUnOwPermP{}    & Route network \\
\bottomrule
\end{tabular}
\end{table}

\section{Discussion}
\label{sec:discussion}

\subsection{The Methods Contribution}

The central finding is methodological. Table~\ref{tab:rayleigh_vs_binom}
shows empirically that the Rayleigh periodogram and the geometric-null
binomial test are anti-correlated across periods in a geographically
clustered monument corpus: Rayleigh significance appears at arbitrary
periods carrying no proximity signal, and the only genuine proximity
signal appears at a period where Rayleigh is non-significant.

This anti-correlation is a predictable consequence of what each test
measures. The Rayleigh test detects unimodal phase concentration: the
case where all sites share one preferred longitude modulo $T$. The
binomial proximity test detects multimodal proximity enrichment: the
case where sites fall near the full set of harmonics regardless of
which specific harmonic they are near. These are different hypotheses.
The one the archaeoastronomical literature has typically meant to test
(``sites cluster near longitude harmonics'') is the latter, and using
Rayleigh to test it introduces a systematic error.

The geometric-null binomial framework resolves this. The geometric null
$p_0 = 2d/T$ is transparent, requires no distributional assumptions,
and correctly treats geographic clustering as neutral. When multiple
independently assembled corpora are available, Fisher combination of
their binomial p-values gives a joint estimate whose significance is
not inflated by within-corpus clustering, because the corpora's
geographic biases are non-overlapping. This framework is applicable
to any spatial periodicity hypothesis of this form regardless of
geographic or cultural context.

\subsection{Interpretation of the Empirical Illustration}

The three-corpus demonstration finds consistent directional excess over
the geometric null at the $3\degr$ / ${\sim}35\degr$E candidate frame
(Fisher $p = \geoFisherAPDisp{}$ at Tier~A). The UNESCO corpus is the most
geographically distributed result, with A-tier sites spanning the full
Eurasian corridor. Null~A shows its enrichment is reproducible from
dome/stupa sites' own longitude geography ($p = \geoNullDomeBootP{}$, ns),
confirming the signal is carried by where hemispherical monuments are
located. Null~C, however, shows that the broader same-footprint UNESCO
background expects only ${\sim}\geoNullDomeRestrictedMeanfive{}$ A$++$ hits
versus the observed \NcircTierApp{} ($p = \geoNullDomeRestrictedPfive{}$),
indicating partial type-specificity within that geography: the dome subset
is more concentrated near harmonics than other monuments occupying the same
longitude footprint. These two nulls point in opposite directions and should
be read together: the enrichment is embedded in hemispherical monument
geography (Null~A), but that geography is itself non-randomly placed relative
to the grid compared with the surrounding UNESCO background (Null~C). The Wikidata
result is substantially carried by the Java/Borobudur node, which is
historically the dominant node of the stupa tradition
\citep{miksic1990borobudur}: the cluster's position on a $3\degr$
harmonic is the signal the geometric-null test is designed to detect.
The OSM result, assembled by a process with no relationship to UNESCO
or Wikidata curation, independently replicates the directional excess
($p = \osmStupaATierBinomP{}$ at Tier~A) and the same anchor
($K = \npcOsmStupaK{}$ with Wikidata).

The present data neither confirm nor rule out any particular
causal interpretation. The statistical pattern is consistent
with several mechanisms that cannot be distinguished with the
available evidence. Non-intentional explanations include geographic
channeling by trade routes and river systems following the Eurasian
corridor, cultural diffusion patterns concentrating stupa
construction at historically dominant nodes \citep{sen2014silk},
and selection effects in the UNESCO inscription process.
An intentional placement interpretation would require pre-modern
continental longitude determination at sub-degree precision, for
which no documentary or archaeological evidence has been identified
\citep{hunger2019mul, neugebauer1975history}; but absence of
evidence is not evidence of absence, and this paper does not
attempt to adjudicate the question. The contribution of the present
work is to establish what the correct statistical test is and what
it finds when applied to three independent corpora --- not to
resolve the mechanism.

\subsection{Limitations}

The analysis is conditional on the stated candidate frame.
An anchor-discovery correction places the Levantine corridor in the
top \anchorSweepTopPct\% globally ($p = \anchorMaxPermP{}$,
$^\dagger$), a boundary condition on phase-free discovery claims. The
type-specificity argument compares hemispherical monuments against the
full UNESCO corpus, which includes non-architectural sites; a
within-footprint comparison against non-hemispherical architectural
monuments would provide a stronger test. Geocoding precision is a
secondary concern: centroid displacement of 20--30~km for serial
nominations is sufficient to cross the A$+$ tier boundary; the
per-site audit \citep{tenenbaum2026audit} should be consulted for
individual assignments.

\section{Conclusion}
\label{sec:conclusion}

The Rayleigh periodogram is structurally misaligned with the hypothesis
it is being used to test in spatial monument studies. ``Sites cluster
near longitude harmonics'' predicts multimodal phase concentration
across multiple harmonics, which produces low Rayleigh $R$. Using
Rayleigh to test it fires on geographic clustering unrelated to
harmonic alignment and misses genuine harmonic alignment. The period-sweep
dissociation in Table~\ref{tab:rayleigh_vs_binom} demonstrates this
in a real monument corpus: Rayleigh significance and binomial proximity
enrichment are anti-correlated across candidate periods.

The geometric-null binomial framework tests the correct hypothesis
directly. Applied to three independently assembled hemispherical
monument corpora at a $3\degr$ / ${\sim}35\degr$E candidate frame,
the framework finds consistent directional excess (Fisher $p = \geoFisherAPDisp{}$
at Tier~A). This demonstration is offered without causal claims: the
mechanism gap is real, conditioning on a discovered frame is
acknowledged, and non-intentional explanations receive equal standing.
The contribution is the framework and the empirical demonstration of
the Rayleigh-binomial dissociation --- not a claim for an ancient
longitude grid.

All data, code, and audit files are openly available at Zenodo
(\citealt{tenenbaum2026audit};
\url{https://doi.org/10.5281/zenodo.19938283}).

\begin{thebibliography}{99}

\bibitem[Crown(1989)]{crown1989}
Crown, A.~D. (1989). \textit{The Samaritans}. J.~C.~B.~Mohr.

\bibitem[Freeman(1976)]{freeman1976}
Freeman, P.~R. (1976). A Bayesian analysis of the megalithic yard.
\textit{JRSS-A}, 139(1), 20--35.

\bibitem[Hunger \& Steele(2019)]{hunger2019mul}
Hunger, H. \& Steele, J. (2019).
\textit{The Babylonian Astronomical Compendium MUL.APIN}. Routledge.

\bibitem[Kendall(1974)]{kendall1974}
Kendall, D.~G. (1974). Hunting quanta.
\textit{Phil.\ Trans.\ R.\ Soc.\ A}, 276, 231--266.

\bibitem[MacDonald(1964)]{macdonald1964}
MacDonald, J. (1964). \textit{The Theology of the Samaritans}. SCM Press.

\bibitem[Magli(2013)]{magli2013}
Magli, G. (2013). \textit{Architecture, Astronomy and Sacred Landscape
in Ancient Egypt}. Cambridge University Press.

\bibitem[Magli(2016)]{magli2016}
Magli, G. (2016). \textit{Archaeoastronomy}. Springer.

\bibitem[Miksic(1990)]{miksic1990borobudur}
Miksic, J.~N. (1990). \textit{Borobudur: Golden Tales of the Buddhas}.
Periplus Editions.

\bibitem[Neugebauer(1975)]{neugebauer1975history}
Neugebauer, O. (1975). \textit{A History of Ancient Mathematical
Astronomy}, 3~vols. Springer.

\bibitem[Purvis(1968)]{purvis1968}
Purvis, J.~D. (1968). \textit{The Samaritan Pentateuch and the Origin
of the Samaritan Sect}. Harvard University Press.

\bibitem[Ruggles(1999)]{ruggles1999astronomy}
Ruggles, C.~L.~N. (1999). \textit{Astronomy in Prehistoric Britain
and Ireland}. Yale University Press.

\bibitem[Ruggles \& Cotte(2010)]{ruggles2010icomos}
Ruggles, C.~L.~N. \& Cotte, M. (Eds.) (2010).
\textit{Heritage Sites of Astronomy and Archaeoastronomy}.
ICOMOS \& IAU.

\bibitem[Ruggles(2015)]{ruggles2015handbook}
Ruggles, C.~L.~N. (Ed.) (2015). \textit{Handbook of Archaeoastronomy
and Ethnoastronomy}. Springer.

\bibitem[Sen(2014)]{sen2014silk}
Sen, T. (2014). The spread of Buddhism. In \textit{The Cambridge World
History}, Vol.~5 (pp.~447--479). Cambridge University Press.

\bibitem[Silva(2020)]{silva2020statistical}
Silva, F. (2020). A probabilistic framework for structural orientations
in skyscape archaeology. \textit{J.\ Archaeol.\ Sci.}, 118, 105138.

\bibitem[Tenenbaum(2026)]{tenenbaum2026audit}
Tenenbaum, S. (2026). \textit{Longitude Quantization in the UNESCO
World Heritage Corpus: Domes, Stupas, and 3\degr\ Harmonics --- Code
and Data} (v1.6.0). Zenodo.
\url{https://doi.org/10.5281/zenodo.19938283}

\bibitem[Thom(1967)]{thom1967}
Thom, A. (1967). \textit{Megalithic Sites in Britain}. Oxford
University Press.

\bibitem[Thom(1971)]{thom1971}
Thom, A. (1971). \textit{Megalithic Lunar Observatories}. Oxford
University Press.

\bibitem[UNESCO World Heritage Centre(2024)]{unescowhc2024xml}
{UNESCO World Heritage Centre}. (2024). \textit{World Heritage List}
[XML dataset]. \url{https://whc.unesco.org/en/list/xml/}.

\bibitem[Williamson \& Bellamy(1983)]{williamson1983leylines}
Williamson, T. \& Bellamy, L. (1983). \textit{Ley Lines in Question}.
World's Work.

\end{thebibliography}

\section*{Data Availability}
All data, code, and audit files: Zenodo (\citealt{tenenbaum2026audit};
\url{https://doi.org/10.5281/zenodo.19938283}).
UNESCO XML: \citealt{unescowhc2024xml}.

\section*{Acknowledgements}
The UNESCO World Heritage Centre maintains the publicly accessible XML
dataset on which this study depends. Research conducted independently.

\section*{Conflict of Interest}
The author declares no competing interests and received no specific
grant from any funding agency.

\section*{Software and AI Disclosure}
Statistical analyses in Python~3 (NumPy, SciPy, pandas). Claude
(Anthropic) assisted with prose drafting; GitHub Copilot (Microsoft)
assisted with code autocompletion. All hypotheses, analytical choices,
and conclusions are solely the author's responsibility.

\end{document}
